---
title: "K-Means from Scratch, in Base R"
subtitle: ""
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 6,
  fig.height = 4.5
)
```

# Introduction

**What the algorithm does?** You give it points and a number `k`; it splits the points into
`k` groups so that points in a group sit close together. No labels, just geometry.

**What it minimizes? `inertia`.** Also called *within-cluster sum of squares* (WCSS): for every
point, take the **squared** distance to its own cluster's center, and add all of them up.
Small inertia = tight clusters. Finding the grouping with the globally smallest inertia is
NP-hard, so K-Means doesn't try. It uses **Lloyd's algorithm**, a heuristic that finds a
*good* answer (a local minimum):

1. **Guess** `k` starting centroids.
2. **Assign** each point to its nearest centroid.
3. **Update** each centroid to the mean of its assigned points.
4. **Repeat** 2-3; **stop** when the centroids stop moving.

Each assign+update cycle can only lower inertia or leave it unchanged (never raise it), and
inertia can't go below 0 — so the loop must settle. It just isn't guaranteed to settle on the
*best* possible answer, which is why initialization matters.

> **Key Takeaway:**
>
> I.) K-Means trades the impossible goal of finding the absolute best clustering for the practical goal of finding a good enough clustering fast. It uses math that always moves downhill, so it's guaranteed to converge—just not to the global optimum.
>
> II.) Always scale your features before K-means. K-means uses Euclidean distance, which is blind to feature ranges—so a feature with range 0-1,000,000 will completely dominate one with range 18-80, causing you to cluster on accident using mostly just that one variable.

# The worked example

Every step below runs on the same 6 points, so you can check numbers by hand.

```{r data}
X <- matrix(c(
  1.0,  2.0,
  1.5,  1.8,
  5.0,  8.0,
  8.0,  8.0,
  1.0,  0.6,
  9.0, 11.0
), ncol = 2, byrow = TRUE)

k <- 2
print(X)
```

For the hand-walkthrough we start the centroids (randomly) on two of the data points themselves:
`(1.0, 2.0)` (row 1) and `(5.0, 8.0)` (row 3).
